Fig2

### 2020-02-21 analysis/02_DESeq2.Rmd
new fig 2

```{r setup}
library(tidyverse)
library(DESeq2)
library(BiocParallel)
register(MulticoreParam(6))
library(caret) 
library(cowplot)

@@

## Warning: This script can take a very long time to run.

DESeq2 was not designed to run on 100+ samples. But, I really like it, so I do it anyways. Some of these commands take like 15 min to run using 6 cores.

@@

>>>>

```{r setup}
library(tidyverse)
library(DESeq2)
library(BiocParallel)
register(MulticoreParam(6))
library(caret)  # for sampling data
library(cowplot)

@@

## Notes

- DESeq2 was not designed to run on 100+ samples. But, I really like it, so I do it anyways. Some of these commands take like 15 min to run using 6 cores. 
_ Also, I don't run every chuck every time. When I want new analysese, I add them to the bottom and set `eval = T` for chunks I don't want to rerurn

@@

## Hypothesis testing - external stimuli and/or prolactin

```{r}
# label each bird with weather or not their pituitary had hi PRL or not
# then make new column with nest, eggs, or chicks, drop controls

PRLdata <- read_csv("../results/PRLvsd.csv") %>%
  mutate(V1 = samples) %>%
  mutate(bird = sapply(strsplit(as.character(samples),'\\_'), "[", 1)) %>%
  select(bird,hiloPRL)

# join with hi lo, and drop controls
colDatahypothesis <- full_join(colData, PRLdata, by = "bird") %>% 
  drop_na() %>%
  select(V1, tissue, sex, sextissue, treatment, hiloPRL) %>% 
  filter(!treatment %in% c("control", "bldg")) %>%
  mutate(external = fct_collapse(treatment, eggs = c("lay", "inc.d3", "inc.d9", "inc.d17"),
                                            chicks = c("hatch", "n5", "n9")))

colDatahypothesis <- as.data.frame(colDatahypothesis)
row.names(colDatahypothesis) <- colDatahypothesis$V1

savecols <- as.character(colDatahypothesis$V1) 
savecols <- as.vector(savecols)

colDatahypothesis %>%
  group_by(sex, tissue, hiloPRL, external) %>%
  summarize(n = n())

hypothesesistesting <- function(whichfactor, up, down, mytissue){

  res <- results(dds, contrast = c(whichfactor, up, down), independentFiltering = T, alpha = 0.1)

  partialfilename = paste("_", down, "_", up, sep = "")
  myfilename = paste0("../results/DESeq2/hypothesis/", mytissue, partialfilename, "_DEGs.csv")

for(i in levels(colDatahypothesis$sextissue)){

  newcolData <- subsetcolData2(colDatahypothesis, i)

  dds <- DESeqDataSetFromMatrix(countData = newcountData,
                                colData = newcolData,
                                design = ~ external + hiloPRL )
  dds <- dds[rowSums(counts(dds) > 1) >= 10]  # filter more than sample with less 0 counts
  print(dds)
  print(dim(dds))
  dds <- DESeq(dds, parallel = TRUE) # Differential expression analysis

  myfilename = paste0("../results/DEseq2/hypothesis/", i, "_vsd.csv")
  write.csv(vsd, myfilename)

  # save differential gene expression results
  # sexmalefemale <- hypothesesistesting("sex", "male", "female",  i) 
  externalschickseggs <- hypothesesistesting("external", "chicks", "eggs",  i) 
  hiloPRL <- hypothesesistesting("hiloPRL", "hi","lo",  i) 
}
```

### 2020-02-21 analysis/figure-2.Rmd
new fig 2

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(kableExtra)
source("../R/themes.R")

@@

```{r allDEGs}

DEG_path <- "../results/DEseq2/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
DEG_files

@@

allDEG$tissue <- factor(allDEG$tissue , levels = tissuelevel)
allDEG$comparison <- factor(allDEG$comparison , levels = comparisonlevels)
allDEG$direction <- factor(allDEG$direction, levels = charlevels)

overlappingDEGs <- allDEG %>%
  filter(tissue == "hypothalamus",
         comparison != "control_bldg") %>%
  mutate(group = paste(sex, tissue,comparison, sep = ".")) %>%
  group_by(gene) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
overlappingDEGs

HYPcandidategenes <- c("OXT", "AVP", "GNRH1",  "AR", "POMC", "AGRP",
                       "CRH", "AVPR1A", "AVPR1B", "AVPR2",
                       "CYP19A1", "DRD1", "DRD2", "PRL", "PRLR")

suppletable1 <- allDEG %>%
  filter(gene %in% HYPcandidategenes, 
         tissue == "hypothalamus") %>%
  group_by(sex, tissue, comparison) %>%
  summarize(genes = str_c(gene, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = genes ) %>%
  select(sex, tissue, control_bldg,  
         inc.d9_inc.d17, hatch_n5)  %>%
  arrange(sex, tissue)
suppletable1

write.csv(suppletable1, "../results/suppltable-1.csv", row.names = F)
```

## variance stabilized gene expression  (vsd)

```{r vsd}
geneids <- read_csv("../metadata/00_geneinfo.csv")

vsd_path <- "../results/DEseq2/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts") 
```

```{r fig2, fig.width=7, fig.height=7}

p1 <- allDEG %>%
  filter(comparison == "control_bldg") %>%
  filter(tissue == "hypothalamus") %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_grid(tissue~sex) +
  theme_B3() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
         #axis.text.x = element_blank(),
        legend.position = "none",
        strip.text.y = element_blank())  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = " ", y = "Hypothalamic DEGs") +
  scale_fill_manual(values = colorscharnew,
                       name = " ",
                       drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black")  +
  ylim(0,4200)

p2 <- allDEG %>%
  filter(comparison != "control_bldg") %>%
    filter(tissue == "hypothalamus") %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_wrap(~sex) +
  theme_B3() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
         #axis.text.x = element_blank(),
        legend.position = "none",
        strip.placement = "right")  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = "Sequential Parental Care Stages  ", y = NULL) +
  scale_fill_manual(values = colorscharnew,
                       name = " ", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black")  +
  ylim(0,1150)

forlegend <- allDEG %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
   scale_fill_manual(values = colorscharnew,
                       name = "increased in", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  guides(color = F) +
  guides(fill= guide_legend(nrow =1)) +
  theme_B3() +
  theme(legend.position = "bottom")

mylegend <- get_legend(forlegend)

barplots <- plot_grid(p1,p2, nrow=1, rel_widths = c(0.9,3))

plottopgenes <- function(whichgenes, whichtissue, whichsex, mysubtitle, whichtstages){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)  %>%
    drop_na()

  candidates$treatment <- factor(candidates$treatment, levels = alllevels)

  p <- candidates %>%
    filter(treatment %in% whichtstages) %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot() + 
    facet_wrap(~gene, scales = "free_y") + 
    theme_B3() +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text = element_text(face = "italic")) +
    labs(y = mysubtitle, x = "Parental stage") 
  return(p)

}

hypDEGs <- c("OXT", "AVP", "GNRH1",  "AR")
topgenes <- plottopgenes(hypDEGs, "hypothalamus", c("female", "male"), "Hypothalamic expression", charlevels)

plot_grid(barplots, topgenes, nrow = 2, rel_heights = c(0.4,0.6))

@@

```{r suppletable1}
kable(suppletable1)
```

```{r writefiles}
#write_csv(allDEG, "../results/DEGs-char.csv")
```

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(kableExtra) 
library(ggtext) # for italic fonts
library(ggsignif) # fot stats on plots

source("../R/themes.R")

@@

```{r allDEGs}
DEG_path <- "../results/DEseq2/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
#DEG_files

@@

allDEG$tissue <- factor(allDEG$tissue , levels = tissuelevel)
allDEG$comparison <- factor(allDEG$comparison , levels = comparisonlevels)
allDEG$direction <- factor(allDEG$direction, levels = charlevels)
```

```{r figurefunctions}

makebargraph <- function(whichtissue, myylab, lowlim, higherlim){
  p <- allDEG %>%
    filter(tissue == whichtissue,
           comparison != "control_bldg") %>%
  ggplot(aes(x = comparison,  fill = direction)) +
    geom_bar(position = "dodge") +
    facet_grid(tissue~sex) +
    theme_B3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")  +
    guides(fill = guide_legend(nrow = 1)) +
    labs(x = "Sequential parental care stage comparisons", 
         y = myylab,
         subtitle = " ") +
  scale_fill_manual(values = allcolors,
                       name = " ",
                       drop = FALSE) +
  scale_color_manual(values = allcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black")  +
  ylim(lowlim, higherlim)
  return(p)
}

@@

## Internal versus external

```{r hypothesis}
DEG_path <- "../results/DEseq2/hypothesis/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
#DEG_files

allDEG2 <- DEG_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name") %>% 
  mutate(DEG = sapply(strsplit(as.character(file_name),'./results/DEseq2/hypothesis/'), "[", 2))  %>% 
  mutate(DEG = sapply(strsplit(as.character(DEG),'_diffexp.csv'), "[", 1))  %>% 
  mutate(tissue = sapply(strsplit(as.character(DEG),'\\.'), "[", 1)) %>%
  mutate(down = sapply(strsplit(as.character(DEG),'\\_'), "[", 3)) %>%
  mutate(up = sapply(strsplit(as.character(DEG),'\\_'), "[", 4)) %>%
  mutate(comparison = paste(down,up, sep = "_")) %>%
  mutate(sex = sapply(strsplit(as.character(sextissue),'\\_'), "[", 1)) %>%
  mutate(tissue = sapply(strsplit(as.character(sextissue),'\\_'), "[", 2)) %>%
dplyr::select(sex,tissue,comparison, direction, gene, lfc, padj, logpadj) 
head(allDEG)

allDEG2$tissue <- factor(allDEG2$tissue, levels = tissuelevel)

allDEG2$comparison <- factor(allDEG2$comparison, levels = c("eggs_chicks", "lo_hi"))
allDEG2 <- allDEG2 %>% mutate(comparison = fct_recode(comparison, "lo vs. hi PRL   " = "lo_hi",
                                                      "eggs vs. chicks" = "eggs_chicks"))
allDEG2$direction <- factor(allDEG2$direction, levels = c("eggs", "chicks", "lo", "hi"))

makenewbargraph <- function(whichtissue, myxlab, whichcomparison, lowlim, higherlim){
  p <- allDEG2 %>%
    filter(tissue == whichtissue,
           comparison == whichcomparison) %>%
  ggplot(aes(x = comparison,  fill = direction)) +
    geom_bar(position = "dodge") +
    facet_grid(tissue~sex) +
    theme_B3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none")  +
    guides(fill = guide_legend(nrow = 1)) +
    labs(x = myxlab, 
         y = "Total DEGs") +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
           position = position_dodge(width = 1),
           size = 2.5, color = "black")  + 
      ylim(lowlim, higherlim)
  return(p)
}

## DEGS

```{r fig2, fig.width=7, fig.height=6}
# hyp
p1 <- makebargraph("hypothalamus","DEGs", 0, 1250) + theme(axis.text.x = element_blank(), 
                                                               axis.title.x = element_blank())

# pit
p2 <- makebargraph("pituitary","DEGs", 0, 1250)  +  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), 
                                                              strip.text.x = element_blank())

# gon
p3 <- makebargraph("gonad","DEGs", 0, 1250) +  theme(strip.text.x = element_blank())

bcd <- plot_grid(p1,p2,p3, nrow = 3, rel_heights = c(1.2,1,1.5), labels = c("b", "c", "d"), label_size = 12)

expdesign <- png::readPNG("../figures/images/DoveParentsRNAseq_DEGS.png")
expdesign <- ggdraw() +  draw_image(expdesign, scale = 1)

plot_grid(expdesign, bcd, labels = c("a", " "), label_size = 12, nrow = 2, rel_heights = c(0.2,1))

b <- makenewbargraph(tissuelevel, "External stimuli", "eggs vs. chicks", 0, 1250)  + theme(strip.text.y = element_blank()) +
  scale_fill_manual(values = c("#80cdc1", "#018571"))
c <- makenewbargraph(tissuelevel, "Pituitary PRL expression   ", "lo vs. hi PRL   ", 0, 1250) + labs(y = NULL) + theme(axis.text.y = element_blank()) +
  scale_fill_manual(values = c("#dfc27d", "#a6611a"))

plot_grid(b, c, rel_widths = c(1,1), nrow = 1, labels = "auto", label_size = 12)
```

## total degs

```{r}
allDEG %>%
  group_by(tissue, comparison) %>%
  summarize(totalDEGs = n()) %>%
  arrange(tissue, comparison)
```

## candidate genes 
```{r writefiles}
candidategenes <- c("OXT", "AVP", "GNRH1",  "AR", "POMC", "AGRP",
                       "CRH", "AVPR1A", "AVPR1B", "AVPR2",
                       "CYP19A1", "DRD1", "DRD2", "PRL", "PRLR", "SOX9")

table1 <- allDEG %>%
  filter(gene %in% candidategenes) %>%
  group_by(sex, tissue, comparison) %>%
  summarize(genes = str_c(gene, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = genes ) %>%
  select(sex, tissue, control_bldg, lay_inc.d3, inc.d3_inc.d9,
        inc.d9_inc.d17, hatch_n5)  %>%
  arrange( tissue, sex)
table1

write_csv(table1, "../results/table1.csv")
```
